Storm surge modelling in the Strait of Georgia

### 2014-12-05 RoughWork/manuscript.tex
Edits from Susan and Kate.

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damaged caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There is also a small contribution to increased sea level due to positive sea surface height anomalies propagating northward from the equator along the eastern Pacific coast during El Nino years \citep{strub2002altimeter}.

@@

The five hindcast simulations have been used to assess the model's skill and to learn more about how the storm surge propagates through the Salish Sea domain.  First, a general assessment of the model's skill is presented in two hindcasts during Feb 2006 and Nov 2006 in section \ref{sec:hind}. The Feb 2006 case is also used to determine the importance of modelling tidal forcing with the surge propagation. Second, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed for the Feb 2006 case in section \ref{sec:factors}. Third, the spatial variability of the surge as it propagates through the domain is examined for a Dec 2006 storm surge in section \ref{sec:spatial}. Fourth, a high wind event that did not result in a large surge or flooding for a Nov 2009 case is discussed in section \ref{sec:wind}. Finally, comparisons between hindcasts using high resolution (HRDPS) and low resolution (CGRF) atmospheric forcing is provided in section \ref{sec:res}.

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not include. Other findings have reported the inverse barometer effect contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}, but only a minor effect is noticeable here.

>>>>

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damage caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There is also a small contribution to increased sea level due to positive sea surface height anomalies propagating northward from the equator along the eastern Pacific coast during El Nino years \citep{strub2002altimeter}.

@@

The five hindcast simulations have been used to assess the model's skill and to learn more about how the storm surge propagates through the Salish Sea domain.  First, a general assessment of the model's skill is presented in two hindcasts during Feb 2006 and Nov 2006 in section \ref{sec:hind}. The Feb 2006 case is also used to determine the importance of modelling tidal forcing with the surge propagation. Second, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed for the Feb 2006 case in section \ref{sec:factors}. Third, the spatial variability of the surge as it propagates through the domain is examined for a Dec 2006 storm surge in section \ref{sec:spatial}. Fourth, a high wind event that did not result in a large surge or flooding for a Nov 2009 case is discussed in section \ref{sec:wind}. Finally, comparisons between hindcasts using high resolution (HRDPS) and low resolution (CGRF) atmospheric forcing is provided for the Dec 2012 case in section \ref{sec:res}.

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not included. Other findings have reported the inverse barometer effect contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}, but only a minor effect is seen here.
